

## Overview
```{r} 

library("dplyr")
library("tidyr")
library("ggplot2")
library("ggthemes")
library("tidyverse")



data <- read.csv("APS - Mod Pred/california.csv", stringsAsFactors = TRUE)
data <- as_tibble(data)


summary(data)
```

## Breaking down dataset features







## Analysis of Median House Value based on features
### Longitude & Latitude

```{r}
library("mapdata") # Supplement to maps package


california <- subset(map_data("state"), region == "california")

ggplot(data = california, mapping = aes(x = long, y = lat)) +
    geom_polygon(colour = "black", fill = "white", alpha = 0.2) +
    geom_point(
        data = data, aes(
            x = longitude, y = latitude, colour = median_house_value
        ),
        alpha = 0.4
    ) +
    scale_colour_continuous() +
    xlab("Longitude") +
    ylab("Latitude") +
    ggtitle("California Map: Median House Price") +
    theme_bw() +
    theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    )
```


## Ocean Proximity

```{r}
library("ggridges")
library("scales")

ggplot(data = california, mapping = aes(x = long, y = lat)) +
    geom_polygon(colour = "black", fill = "white", alpha = 0.2) +
    geom_point(data = data, aes(x = longitude, y = latitude, color = ocean_proximity), alpha = 0.4, size = 2) +
    scale_color_viridis_d() +
    xlab("Longitude") +
    ylab("Latitude") +
    ggtitle("Map of California: Ocean Proximity") +
    theme_bw() +
    theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    )


ggplot(data, aes(x = median_house_value, y = ocean_proximity, fill = ocean_proximity)) +
    geom_density_ridges_gradient() +
    scale_fill_brewer() +
    theme_bw() +
    labs(title = "Median House Price by Ocean Proximity", x = "Median house value (in thousands)", y = "Ocean proximity") +
    scale_x_continuous(
        breaks = round(seq(min(data$median_house_value), max(data$median_house_value), length.out = 8), 3),
        labels = scales::dollar_format(scale = .001, prefix = "$")
    ) +
    theme(
        legend.position = "none",
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    )
```





# Feature engineering



```{r}
cali_cities <- read.csv(
    "APS - Mod Pred/cal_cities_lat_long.csv",
    stringsAsFactors = TRUE
)
cali_pop <- read.csv(
    "APS - Mod Pred/cal_populations_city.csv",
    stringsAsFactors = TRUE
)
california_cities <- as_tibble(merge(cali_cities, cali_pop, by.x = "Name", by.y = "City"))


summary(california_cities)
```


```{r}
ggplot(data = california, mapping = aes(x = long, y = lat)) +
    geom_polygon(colour = "black", fill = "white", alpha = 0.2) +
    geom_point(
        data = california_cities, aes(
            x = Longitude, y = Latitude, size = pop_april_1990
        ),
        alpha = 0.4
    ) +
    scale_size_area(max_size = 25) +
    xlab("Longitude") +
    ylab("Latitude") +
    ggtitle("California Map: Cities Population 1990") +
    theme_bw() +
    theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    )
```


### Appending County & City to the dataset
```{r} 


knn_dist <- as_tibble(FNN::get.knnx(
    select(california_cities, "Longitude", "Latitude"),
    select(data, "longitude", "latitude"),
    k = 1, algorithm = c(
        "kd_tree", "cover_tree",
        "CR", "brute"
    )
)) %>% select(nn.index)

knn_dist <- as_tibble(as.data.frame(knn_dist$nn.index))


data$city <- california_cities$Name[match(knn_dist$V1, rownames(california_cities))]

data <- merge(data,
    california_cities %>%
        select(Name, pop_april_1990, County),
    by.x = "city", by.y = "Name"
) %>%
    as_tibble(.)
```

### County impact
```{r}


california_counties <- subset(map_data("county"), region == "california")
california_counties$subregion <- as.character(california_counties$subregion)

county_labels <- california_counties %>%
    group_by(subregion) %>%
    summarize_at(vars(long, lat), ~ mean(range(.)))


data %>%
    group_by(County) %>%
    summarise_at(vars(median_house_value), funs(mean)) %>%
    ungroup() %>%
    rename_at("County", ~"subregion") %>%
    mutate(subregion = as.character(tolower(subregion))) %>%
    inner_join(california_counties, .) %>%
    mutate(subregion = stringr::str_to_sentence(subregion)) -> california_counties


ggplot(data = california, mapping = aes(x = long, y = lat)) +
    coord_fixed(1.3) +
    geom_polygon(color = "black", fill = NA) +
    geom_polygon(data = california_counties, color = "black", aes(fill = median_house_value, group = group)) +
    scale_fill_gradient(trans = "log10") +
    labs(title = "Median House Value by County", x = "Longitude", y = "Latitude") +
    theme_bw() +
    theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    ) +
    geom_text(data = county_labels, aes(label = stringr::str_to_title(subregion), x = long, y = lat), color = "white", size = 3.5, fontface = "bold")
```


### Population 
* Block Population
* City Population
* 300km Radius Population

```{r}


knn_radius <- as_tibble(FNN::get.knnx(
    select(california_cities, "Longitude", "Latitude"),
    select(data, "longitude", "latitude"),
    k = 10, algorithm = c(
        "kd_tree", "cover_tree",
        "CR", "brute"
    )
))

as_tibble(as.data.frame(knn_radius$nn.dist)) -> knn_radius_dist
as_tibble(as.data.frame(knn_radius$nn.index)) -> knn_radius_index


knn_radius_dist[knn_radius_dist > 0.3] <- NA
knn_radius_index[is.na(knn_radius_dist)] <- NA


california_cities %>%
    mutate(., rownumber = row_number()) %>%
    select(rownumber, pop_april_1990) -> a

data$radius_pop <- lapply(knn_radius_index, function(x) a$pop_april_1990[match(x, a$rownumber)]) %>%
    as.data.frame(.) %>%
    rowSums(., na.rm = TRUE)

library(gridExtra)


p1 <- ggplot(data, aes(x = population, y = median_house_value)) +
    stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white", bins = 4)

p2 <- ggplot(data, aes(x = pop_april_1990, y = median_house_value)) +
    stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white", bins = 4)

p3 <- ggplot(data, aes(x = radius_pop, y = median_house_value)) +
    stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white", bins = 4)

plist <- list(p1, p2, p3)

grid.arrange(grobs = plist, top = "Population and Median House Price", ncol = 3)
```


## Temperature & Precipitation

```{r}
library("openxlsx")
library("cowplot")

precipitation <- as.data.frame(read.xlsx("APS - Mod Pred/4-pcp-199004-1.xlsx")) %>%
    select(Location, Value) %>%
    rename(., mean_prec = Value)

temp <- as.data.frame(read.xlsx("APS - Mod Pred/4-tavg-199004-1.xlsx")) %>%
    select(Location, Value) %>%
    rename(., mean_temp = Value)


climate <- merge(temp, precipitation, by = "Location") %>%
    mutate_at(vars(Location), function(x) gsub("County", "", x)) %>%
    mutate_at(vars(Location), function(x) gsub("^ +| +$|( ) +", "\\1", x)) %>%
    mutate_at(vars(mean_temp, mean_prec), function(x) as.double(x))




# create 3 buckets
quantiles_temp <- climate %>%
    pull(mean_temp) %>%
    quantile(probs = seq(0, 1, length.out = 4))

quantiles_prec <- climate %>%
    pull(mean_prec) %>%
    quantile(probs = seq(0, 1, length.out = 4))



# create color scale that encodes two variables
# red for temperature and blue for mean precipitation
# the special notation with gather is due to readibility reasons
bivariate_color_scale <- tibble(
    "3 - 3" = "#3F2949", # high temp , high precipitation
    "2 - 3" = "#435786",
    "1 - 3" = "#4885C1", # low temp, high precipitation
    "3 - 2" = "#77324C",
    "2 - 2" = "#806A8A", # medium temp, medium precipitation
    "1 - 2" = "#89A1C8",
    "3 - 1" = "#AE3A4E", # high temp, low precipitation
    "2 - 1" = "#BC7C8F",
    "1 - 1" = "#CABED0" # low temp, low precipitation
) %>%
    gather("group", "fill")




climate %>%
    mutate(
        temp_quantiles = cut(
            mean_temp,
            breaks = quantiles_temp,
            include.lowest = TRUE
        ),
        prec_quantiles = cut(
            mean_prec,
            breaks = quantiles_prec,
            include.lowest = TRUE
        ),
        # by pasting the factors together as numbers we match the groups defined
        # in the tibble bivariate_color_scale
        group = paste(
            as.numeric(temp_quantiles), "-",
            as.numeric(prec_quantiles)
        )
    ) %>%
    # we now join the actual hex values per "group"
    # so each municipality knows its hex value based on the his gini and avg
    # income value
    left_join(bivariate_color_scale, by = "group") %>%
    rename_at("Location", ~"subregion") %>%
    mutate(subregion = as.character(tolower(subregion))) %>%
    merge(california_counties, ., by.x = "subregion", by.y = "subregion") %>%
    as_tibble(.) -> climate



bivariate_color_scale %>%
    separate(group, into = c("temp", "prec"), sep = " - ") %>%
    mutate(
        temp = as.integer(temp),
        prec = as.integer(prec)
    ) -> bivariate_color_scale


legend <- ggplot() +
    geom_tile(
        data = bivariate_color_scale,
        mapping = aes(
            x = temp,
            y = prec,
            fill = fill
        )
    ) +
    scale_fill_identity() +
    labs(
        x = "Higher Temperature ⟶️",
        y = "Higher Precipitation ⟶️"
    ) +
    theme_map() +
    # make font small enough
    theme(
        axis.title = element_text(size = 12),
        axis.title.y = element_text(angle = 90, vjust = 0.5)
    ) +
    # quadratic tiles
    coord_fixed()

graph <- ggplot(data = california, mapping = aes(x = long, y = lat)) +
    coord_fixed(1.3) +
    geom_polygon(color = "black", fill = NA) +
    geom_polygon(data = climate, color = "black", aes(fill = fill, group = group.x)) +
    scale_fill_identity() +
    labs(title = "County Temperature and Precipitation Bivariate", x = "Longitude", y = "Latitude") +
    theme_bw() +
    theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    ) +
    geom_text(data = county_labels, aes(label = stringr::str_to_title(subregion), x = long, y = lat), color = "black", size = 4, fontface = "bold")

ggdraw() +
    draw_plot(graph, 0, 0, 1, 1) +
    draw_plot(legend, 0.02, 0.075, 0.2, 0.2)
```

```{r}

data <- merge(data %>% mutate(County = as.character(County)), climate, by.x = "County", by.y = "Location")

p1 <- data %>%
    group_by(mean_temp) %>%
    summarise(mean_temp_median_house_value = mean(median_house_value)) %>%
    ggplot(., aes(x = mean_temp, mean_temp_median_house_value)) +
    geom_point() +
    geom_line() +
    geom_smooth(method = "loess") +
    theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    ) +
    labs(title = "House Values by Mean Temperature 1990", x = "Temperature in April 1990", y = "House Values") +
    theme_bw()

p2 <- data %>%
    group_by(mean_prec) %>%
    summarise(mean_prec_median_house_value = mean(median_house_value)) %>%
    ggplot(., aes(x = mean_prec, mean_prec_median_house_value)) +
    geom_point() +
    geom_line() +
    geom_smooth(method = "loess") +
    theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    ) +
    labs(title = "House Values by Mean Precipitation 1990 ", x = "Precipitation in April 1990", y = "House Values") +
    theme_bw()



plist2 <- list(p1, p2)
grid.arrange(grobs = plist2, top = "Climate and Median House Value")
```


```{r}

library(viridis)

data %>%
group_by(mean_prec, mean_temp)%>%
summarise(mean= mean(median_house_value), households=mean(households)) %>% 
  arrange(desc(mean_prec)) %>%
  ggplot(aes(x=mean_prec, y=mean_temp, size=mean,color=households)) +
    geom_point(alpha=0.6) +
    scale_size(range = c(.1, 30), name="Mean House Value") + 
    theme_bw() +
    labs(title="House Values, Climate and Number of Households", x='Mean Precipitation 1990', y='Mean Temperature 1990') + theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    )

```



## BEDROOMS & ROOMS 
```{r}

data %>%
    mutate(total_bedrooms_cat = cut(total_bedrooms,
        breaks = c(seq(min(data$total_bedrooms), max(data$total_bedrooms), length.out = 10)),
        dig.lab = 5
    )) %>%
    mutate(
        total_rooms_cat =
            cut(total_rooms,
                breaks = c(seq(min(data$total_rooms), max(data$total_rooms), length.out = 10)),
                dig.lab = 5
            )
    ) %>%
    drop_na(.) -> data


p1 <- ggplot(data, aes(x = total_bedrooms_cat, y = median_house_value)) +
    geom_boxplot(fill = "#3182bd") +
    theme_bw() +
    labs(x = "Total Number of Bedrooms", y = "Median House Price")



p2 <- ggplot(data, aes(x = total_rooms_cat, y = median_house_value)) +
    geom_boxplot(fill = "#3182bd") +
    theme_bw() +
    labs(x = "Total Number of Rooms ", y = "Median House Price")



plist <- list(p1, p2)
grid.arrange(grobs = plist, cols = 2, top = "Bedrooms and Rooms in block and Median House Value")
```



```{r} 

data %>%
    mutate(categorical_age = cut(
        housing_median_age,
        breaks = c(seq(min(housing_median_age), max(housing_median_age), 4))
    )) %>%
    group_by(categorical_age) %>%
    summarise(sd = sd(median_house_value, na.rm = TRUE), median_house_value = mean(median_house_value), ) %>%
    ggplot(., aes(x = categorical_age, y = median_house_value, fill = "black")) +
    geom_bar(stat = "identity", fill = "#3182bd") +
    geom_errorbar(aes(ymin = median_house_value - sd, max = median_house_value + sd), width = .2, position = position_dodge(width = 0.7)) +
    theme_bw() +
    theme(
        panel.spacing = unit(0.1, "lines"),
        strip.text.x = element_text(size = 8),
        plot.title = element_text(face = "bold")
    ) +
    labs(x = "Building age brackets", y = "Median House Price", title = "Age of buildings in block and Median House Prices")

```


<div style='padding:10px; background:lightgreen;border-radius:5px'>
    <b>Insights:</b><br />
    - Most of the houses in this dataset has only 1 floor.<br />
    - All the houses in this dataset has 1 to 3 floors. <br />
    - The prices of the majority of houses with 2 floors are higher than those of 1 or 3 floors. <br />
    - The houses with only 1 floor has the lowest average price.
</div>



# TODO : SOCIOECONOMICAL VARIABLES
# TODO : DISTANCE TO OCEAN
